Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-07-11 paper-new/discussion.tex
clean up intro and move section on prior approaches to discussion

>>>>

\eric{I moved this from the intro. We may need to revise it.}

\subsection{Limitations of Prior Work}

We give an overview of current approaches to proving graduality of
languages and why they do not meet our criteria of a reusable semantic
framework.

\subsubsection{From Static to Gradual}

Current approaches to constructing languages that satisfy the
graduality property include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
\cite{cimini-siek2016}.  These allow the language developer to start
with a statically typed language and derive a gradually typed language
that satisfies the gradual guarantee. The main downside to these
approaches lies in their inflexibility: since the process in entirely
mechanical, the language designer must adhere to the predefined
framework.  Many gradually typed languages do not fit into either
framework, e.g., Typed Racket \cite{tobin-hochstadt06,
  tobin-hochstadt08} and the semantics produced is not always the
desired one.
%
Furthermore, while these frameworks do prove graduality of the
resulting languages, they do not show the correctness of the
equational theory, which is equally important to sound gradual typing.

\subsubsection{Double Categorical Semantics}

New and Licata \cite{new-licata18} developed an axiomatic account of
the graduality relation on a call-by-name cast calculus terms and
showed that the graduality proof could be modeled using semantics in
certain kinds of \emph{double categories}, categories internal to the
category of categories. A double category extends a category with a
second notion of morphism, often a notion of ``relation'' to be paired
with the notion of functional morphism, as well as a notion of
functional morphisms preserving relations. In gradual typing the
notion of relation models type precision and the squares model the
term precision relation. This approach was influenced by the semantics
of parametricity using reflexive graph categories
\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
categories are essentially double categories without a notion of
relational composition. In addition to capturing the notions of type
and term precision, the double categorical approach allows for a
\emph{universal property} for casts: upcasts are the \emph{universal}
way to turn a relation arrow into a function in a forward direction
and downcasts are the dual universal arrow.  Later, New, Licata and
Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
call-by-name to call-by-value as well by giving an axiomatic theory of
type and term precision in call-by-push-value. This left implicit any
connection to a ``double call-by-push-value'', which we make explicit
in Section~\ref{sec:cbpv}.

With this notion of abstract categorical model in hand, denotational
semantics is then the work of constructing concrete models that
exhibit the categorical construction. New and Licata
\cite{new-licata18} present such a model using categories of
$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
Maillard, Tabareau and Tanter to prove graduality of a gradual
dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
domain-theoretic approach meets our criteria of being a semantic
framework for proving graduality, but suffers from the limitations of
classical domain theory: the inability to model viciously
self-referential structures such as higher-order extensible state and
similar features such as runtime-extensible dynamic types. Since these
features are quite common in dynamically typed languages, we seek a
new denotational framework that can model these type system features.

The standard alternative to domain theory that scales to essentially
arbitrary self-referential definitions is \emph{step-indexing} or its
synthetic form of \emph{guarded recursion}. A series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions.

### 2024-07-11 paper-new/intro.tex
clean up intro and move section on prior approaches to discussion

The main contribution of this work is a compositional denotational
semantics for gradually typed languages that validates $\beta\eta$
equality and satisfies a graduality theorem. A great deal of the work
has further been verified in guarded cubical Agda, demonstrating that
the semantics is readily mechanizable.

@@

The paper is laid out as follows:
\begin{enumerate}
\item In Section\ref{sec:syntax} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section\ref{sec:concrete-terms} we develop a first
  denotational semantics in synthetic guarded domain theory that
  satisfies adequacy and respects the equational theory, but does not
  validate graduality. We use this to introduce some of our main
  technical tools: modeling recursive types in guarded type theory and
  modeling effects using call-by-push-value.
\item In Section\ref{sec:relational-issues} we show where the
  New-Licata classical domain theoretic approach fails to adapt
  cleanly to the guarded setting and explore the difficulties of
  proving graduality in an intensional model.
\item In Section\ref{sec:relational-semantics} we
\item In Section\ref{sec:discussion} we discuss prior work on proving
  graduality, our partial mechanization of these results and discuss
  future directions for denotational semantics of gradual typing.
\end{enumerate}

\subsection{Limitations of Prior Work}

We give an overview of current approaches to proving graduality of
languages and why they do not meet our criteria of a reusable semantic
framework.

\subsubsection{From Static to Gradual}

Current approaches to constructing languages that satisfy the
graduality property include the methods of Abstracting Gradual Typing
\cite{garcia-clark-tanter2016} and the formal tools of the Gradualizer
\cite{cimini-siek2016}.  These allow the language developer to start
with a statically typed language and derive a gradually typed language
that satisfies the gradual guarantee. The main downside to these
approaches lies in their inflexibility: since the process in entirely
mechanical, the language designer must adhere to the predefined
framework.  Many gradually typed languages do not fit into either
framework, e.g., Typed Racket \cite{tobin-hochstadt06,
  tobin-hochstadt08} and the semantics produced is not always the
desired one.
%
Furthermore, while these frameworks do prove graduality of the
resulting languages, they do not show the correctness of the
equational theory, which is equally important to sound gradual typing.

%% For example, programmers often refactor their code, and in so doing they rely
%% implicitly on the validity of the laws in the equational theory.
%% Similarly, correctness of compiler optimizations rests on the validity of the
%% corresponding equations from the equational theory. It is therefore important
%% that the languages that claim to be gradually typed have provably correct
%% equational theories.

% The approaches are too inflexible... the fact that the resulting semantics are too lazy
% is a consequence of that inflexibility.
% The validity of the equational theory captures the programmer's intuitive thinking when they refactor their code

%The downside is that
%not all gradually typed languages can be derived from these frameworks, and moreover, in both
%approaches the semantics is derived from the static type system as opposed to the alternative
%in which the semantics determines the type checking. Without a clear semantic interpretation of type
%dynamism, it becomes difficult to extend these techniques to new language features such as polymorphism.

% [Eric] I moved the next two paragraphs from the technical background section
% to here in the intro.
\subsubsection{Double Categorical Semantics}

New and Licata \cite{new-licata18} developed an axiomatic account of
the graduality relation on a call-by-name cast calculus terms and
showed that the graduality proof could be modeled using semantics in
certain kinds of \emph{double categories}, categories internal to the
category of categories. A double category extends a category with a
second notion of morphism, often a notion of ``relation'' to be paired
with the notion of functional morphism, as well as a notion of
functional morphisms preserving relations. In gradual typing the
notion of relation models type precision and the squares model the
term precision relation. This approach was influenced by the semantics
of parametricity using reflexive graph categories
\cite{ma-reynolds,dunphythesis,reynoldsprogramme}: reflexive graph
categories are essentially double categories without a notion of
relational composition. In addition to capturing the notions of type
and term precision, the double categorical approach allows for a
\emph{universal property} for casts: upcasts are the \emph{universal}
way to turn a relation arrow into a function in a forward direction
and downcasts are the dual universal arrow.  Later, New, Licata and
Ahmed \cite{new-licata-ahmed2019} extended this axiomatic treatment from
call-by-name to call-by-value as well by giving an axiomatic theory of
type and term precision in call-by-push-value. This left implicit any
connection to a ``double call-by-push-value'', which we make explicit
in Section~\ref{sec:cbpv}.

With this notion of abstract categorical model in hand, denotational
semantics is then the work of constructing concrete models that
exhibit the categorical construction. New and Licata
\cite{new-licata18} present such a model using categories of
$\omega$-CPOs, and this model was extended by Lennon-Bertrand,
Maillard, Tabareau and Tanter to prove graduality of a gradual
dependently typed calculus $\textrm{CastCIC}^{\mathcal G}$. This
domain-theoretic approach meets our criteria of being a semantic
framework for proving graduality, but suffers from the limitations of
classical domain theory: the inability to model viciously
self-referential structures such as higher-order extensible state and
similar features such as runtime-extensible dynamic types. Since these
features are quite common in dynamically typed languages, we seek a
new denotational framework that can model these type system features.

The standard alternative to domain theory that scales to essentially
arbitrary self-referential definitions is \emph{step-indexing} or its
synthetic form of \emph{guarded recursion}. A series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these explicit
step-indexed to make formalization of realistic gradual languages
tractible.

% Alternative phrasing:
\begin{comment}
\subsubsection{Embedding-Projection Pairs}

The series of works by New, Licata, and Ahmed \cite{new-licata18,new-ahmed2018,new-licata-ahmed2019}
develop an axiomatic account of gradual typing involving \emph{embedding-projection pairs}.
This allows for a particularly elegant formulation of the gradual guarantee.
Moreover, their axiomatic account of program equivalence allows for type-based reasoning about program equivalences.

In \cite{new-licata18}, New and Licata construct a denotational model of the axioms of
gradual typing using tools from classical domain theory. The benefit to this approach is that it is a reusable mathematical theory:
general semantic theorems about gradual domains can be developed independent of any particular syntax and then reused in many different denotational models.
Unfortunately, however, it is unclear how to extend these domain-theoretic techniques to incorporate more advanced language features such as higher-order store, a standard feature of realistic gradually typed languages such as Typed
Racket.
Thus, if we want a reusable mathematical theory of gradual typing that can scale to realistic programming languages, we need to look to different techniques.

In \cite{new-ahmed2018,new-licata-ahmed2019}, a more traditional operational semantics for gradual typing is derived from the axioms.
% The axioms are then proven to be sound with respect to this operational semantics by constructing a logical relation.
A logical relations model is constructed and used to prove that the axioms are sound with respect to the operational semantics.
The downside of this approach is that each new language requires a different logical relation
to prove graduality, even though the logical relations for different languages end up sharing many commonalities.
Furthermore, the logical relations tend to be quite complicated due to a technical requirement known as \emph{step-indexing},
where the stepping behavior of terms must be accounted for in the logical relation.
As a result, developments using this approach tend to require vast effort, with the
corresponding technical reports having 50+ pages of proofs.

% In this approach, a logical relation is constructed and used to show that the equational theory
% is sound with respect to the operational semantics.

% Additionally, while the axioms of gradual type theory are compositional at a ``global'' level,
% they do not compose in the step-indexed setting. One of the main goals of the present work
% is to formulate a composable theory of gradual typing in a setting where the stepping behavior
% is tracked.
\end{comment}

An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain
theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}, abbreviated as
SGDT. The techniques of SGDT allow us to internalize the step-indexed reasoning
normally required in logical relations proofs of graduality, ultimately allowing
us to specify the logical relation in a manner that looks nearly identical to a
typical, non-step-indexed logical relation. In fact, guarded domain theory goes
further, allowing us to define step-indexed \emph{denotational semantics} not
just step-indexed relations, just as easily as constructing an ordinary
set-theoretic semantics.

\eric{This is out of date: the mechanization is now an artifact of the current paper.}
Our longer-term goal is to mechanize these proofs in a reusable way in the
Guarded Cubical Agda \cite{veltri-vezzosi2020} proof assistant, thereby
providing a framework to more easily and conveniently prove that existing
languages satisfy graduality and have sound equational theories.

Moreover, the aim is for designers of new languages to utilize the framework to
facilitate the design of new provably-correct gradually-typed languages with
more complex features.

The main contribution of this work is a categorical and denotational semantics
for gradually typed langauges that models not just the term language but the
graduality property as well.
\begin{enumerate}
\item First, we give a simple abstract categorical model of Gradual Type Theory
using CBPV double categories.
\item Next, we modify this semantics to develop reflexive graph- and double
  categorical models that abstract over the details of step-indexed models, and
  provide a method for constructing such models.
\item We instantiate the abstract construction to provide a concrete semantics
  in informal guarded type theory.
\item We prove that the resulting denotational model is \emph{adequate} for the
  graduality property: a closed term precision $M \ltdyn N : \nat$ has the
  expected semantics, that $M$ errors or $M$ and $N$ have the same extensional
  behavior.
\end{enumerate}

% TODO is it okay to refer to our cast calculus as *the* gradually-typed lambda
% calculus?
\eric{Where should we put the description of the big-step semantics and adequacy
of the semantics? I can put it in the outline of the remainder of the paper, or
I can postpone it until the subsequent section on the syntax of gradual typing.
Adequacy seems to be better introduced there since at that point we have
introduced the notion of term precision.}

The remainder of the paper is organized as follows. In Section
\ref{sec:technical-background}, we provide technical background on synthetic
guarded domain theory and some important details about the specific variant of
SGDT we employ as our ambient language.
%
In Section \ref{sec:GTLC}, we describe the syntactic theory of a standard cast
calculus for gradual typing, which we also refer to as the \emph{gradually-typed
lambda calculus}. We do this to establish a common syntax that will be
interpreted into the models we subsequently construct.
%
In Section \ref{sec:cbpv}, we develop \emph{abstract} categorical
semantics of gradually typed languages with the goal of organizing our
construction of denotational models of gradual typing.
%
In Sections \ref{sec:concrete-term-model} and
\ref{sec:towards-relational-model}, we set out to instantiate the abstract
categorical semantics by constructing a concrete model in a suitable double
category. First, in Section \ref{sec:concrete-term-model}, we give a concrete
model for the \emph{terms} of the gradually-typed lambda calculus. This involves
using the tools of SGDT to solve a domain equation to give the semantics of the
dynamic type. We explain how to \emph{globalize} this model to obtain a model in
the usual (non-guarded) dependent type theory. In the end, we obtain a
\emph{big-step semantics} for closed terms of type $\nat$:
%
\[ \sem{\cdot} \colon (\cdot \vdash M : \nat) \to (\mathbb{N} \to (\sem{\nat} + 1)) \] 
%
where a closed term $M$ of type $\nat$ denotes a \emph{function} from
$\mathbb{N}$ to $\sem{\nat} + 1$ which given input $i$ returns $\inl\, m$ if $M$
terminates with value $m$ in $i$ or fewer steps, and returns $\inr\, *$ if $M$
fails to terminate in $i$ or fewer steps. This is already a significant result,
as it serves to verify the $\beta$ and $\eta$ laws of the equational theory.
%
Then in Section \ref{sec:towards-relational-model}, we attempt to extend the
methodology of the previous section to give a denotational semantics that models
the relational theory of the gradually-typed lambda calculus in addition to the
terms. Interesting issues arise here arising from the \emph{intensional} nature
of SGDT, that is, the fact that \emph{steps are observable}. We then develop
novel solutions to address these issues. We do not complete the construction of
the concrete relational model in this section. Instead, we choose to reformulate
the definition of abstract categorical model to take into account the solutions
to the issues discussed here.

% Interesting issues arise here, and the novel solutions we employ to
% address these issues inform the definition of the abstract models introduced in
% the subsequent section.

With the lessons from the concrete setting at our disposal, in Section
\ref{sec:abstract-models}, we describe revised categorical models of gradual
typing. We actually introduce two separate notions of model: \emph{intensional},
where the stepping behavior of terms is observable, and \emph{extensional},
where we consider terms that differ only in their stepping behavior to be equal.
The extensional model is what we will ultimately use to interpret the syntax and
relational axioms of gradually-typed languages. The principal benefit of the
notion of intensional model is that it admits more compositional reasoning. It
therefore serves as a convenient intermediate setting that enjoys many of the
benefits of the usual double-categorical models of gradual typing.
% We give the definition of intensional model as a sequence of steps each
% building on the previous. 
We then outline how to build an extensional model in a sequence of steps beginning
with a significantly simpler ``input'' model. The construction proceeds in
steps, first building up an intensional model, and then from there extracting an
extensional model.
%We are able to leverage the compositional reasoning afforded by the intensional
%model to more easily meet the requirements of an extensional model.
This gives us a more tractable way to construct models.
Moreover, the abstract definitions and constructions in the definitions make no
mention of SGDT, so that the places where SGDT techniques are employed are
conveniently limited to the data passed as input to the model construction.
%
%In addition to organizing and facilitating the construction of a model, the abstract constructions are described independently of SGDT
%
%we also isolate the parts of the definition the notion of model without any explicit mention of SGDT.
%
Then in Section \ref{sec:concrete-model} we return to the task of defining a
concrete model. Equipped now with the abstract constructions of the previous
section, we provide only the data needed as an input to the construction. Many
of the relevant definitions have already been introduced in Section
\ref{sec:towards-relational-model} and can be provided to the construction
as-is; the remainder of the necessary data is defined in this section.
%
We end the section by sketching a proof of the \emph{adequacy} of our concrete
model with respect to graduality.
%
Finally, in Section \ref{sec:discussion} we conclude with discussion: comparison
to related work, the mechanization of our concrete model in Guarded Cubical
Agda, and directions for future research.

%% \subsection{Outline of Remainder of Paper}

%% \max{this is entirely out of date, update later}

%% % In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
%% % calculus and the graduality theorem.
%% %
%% In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
%% % 
%% In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
%% for which we will prove graduality. We give a semantics to the terms using
%% synthetic guarded domain theory.
%% % Important here are the notions of syntactic
%% % type precision and term precision. For reasons we describe below, we
%% % introduce two related calculi, one in which the stepping behavior of terms is
%% % implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
%% % is made explicit (an \emph{intensional} calculus, $\intlc$).
%% %

%% In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
%% the gradually-typed lambda calculus and describe our approach to assigning a
%% denotational semantics to this ordering.
%% This approach builds on the semantics constructed in the previous section,
%% but extending it to the term ordering presents new challenges.

%% % In Section \ref{sec:domain-theory}, we define several fundamental constructions
%% % internal to SGDT that will be needed when we give a denotational semantics to
%% % the gradual lambda calculus.
%% %This includes the Lift monad, predomains and error domains.
%% %
%% % In Section \ref{sec:semantics}, we define the denotational semantics for the
%% % intensional gradually-typed lambda calculus using the domain theoretic
%% % constructions defined in the previous section.
%% %
%% In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
%% gradual lambda calculus.
%% %
%% In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
%% to the traditional step-indexing approach, as well as possibilities for future work.

%% \subsection{Overview of Results}\label{sec:overview}

%% % This section used to be part of the intro.
%% % \subsection{Proving Graduality in SGDT}
%% % TODO This section should probably be moved to after the relevant background has been introduced?

%% % TODO introduce the idea of cast calculus and explicit casts?

%% In this paper, we will utilize SGDT techniques to prove graduality for a particularly
%% simple gradually-typed cast calculus, the gradually-typed lambda calculus.
%% Before stating the graduality theorem, we introduce some definitions related to gradual typing.

%% % Cast calculi
%% % TODO move this to earlier?
%% Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
%% that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
%% In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
%% The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.

%% % Up and down casts
%% In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
%% $A$ is a \emph{more precise} type than $B$.
%% There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
%% The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
%% %
%% If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
%% and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
%% Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
%% Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.

%% % Syntactic term precision
%% We also have a notion of \emph{syntactic term precision}.
%% If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
%% $M \ltdyn N : A \ltdyn B$ to mean that
%% $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
%% ``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
%% behave the same except that $M$ may error more.
%% Term precision is defined by a set of axioms that capture this intuitive notion.
%% The specific rules for term precision depend on the specifics of the language, but
%% many of the rules can be derived from the typing rules in a straightforward way.
%% % Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
%% % and since
%% Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.

%% % GTLC
%% The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
%% type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
%% between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
%% Section \ref{sec:gtlc-precision}.

%% With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.

%% % \begin{theorem}[Graduality]
%% %   If $M \ltdyn N : \nat$, then either:
%% %   \begin{enumerate}
%% %     \item $M = \mho$
%% %     \item $M = N = \zro$
%% %     \item $M = N = \suc n$ for some $n$
%% %     \item $M$ and $N$ diverge
%% %   \end{enumerate}
%% % \end{theorem}

%% \begin{theorem}[Graduality]
%%   Let $\cdot \vdash M \ltdyn N : \nat$. 
%%   If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
%% \end{theorem}

%% Note:
%% \begin{itemize}

%%   \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
%%   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

%%   \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
%%   $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
%% \end{itemize}

%% % We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.

%% To prove graduality and validate the equational theory, we construct a model of the types
%% and terms and show that all of the axioms for term precision and for equality of terms
%% hold in this model. Modeling the dynamic type presents a challenge in the presence of a
%% language with functions: we want the dynamic type to represent a sum of all possible types
%% in the language, so we write down an recursive equation that the semantic object modeling
%% dynamic type should satisfy. When the language includes function types, this equation involves a
%% negative occurrence of the variable for which we are solving, and so the equation 
%% does not have inductive or coinductive solutions.
%% %
%% To model the dynamic type, we therefore use guarded recursion to define a suitable
%% semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
%% The key is that we do not actually get an exact solution to the equation in the style
%% of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
%% %
%% That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
%% we guard the negative occurrences of the variable by a special operator that
%% specifies that its argument is available ``later''.
%% %This can be seen as a logic that internalizes the notion of step-indexing.
%% See Section \ref{sec:technical-background} for more details on guarded type theory.

%% At a high level, the key parts of our proof are as follows:

%% % TODO revise this
%% \begin{itemize}
%%   \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
%%   or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
%%   computation steps taken by a term are made explicit.
%%   The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
%%   \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
%%   denoted $\extlc$.

%%   \item We define a translation from the surface syntax to the intensional syntax, and
%%   prove a theorem relating the term precision in the surface to term precision in the
%%   intensional syntax.

%%   \item We define a semantics for the intensional syntax in guarded type theory, for both the
%%   terms and for the term precision ordering $\ltdyn$.

%% \end{itemize}

>>>>

The main contribution of this work is a compositional denotational
semantics for gradually typed languages that validates $\beta\eta$
equality and satisfies a graduality theorem. A great deal of the work
has further been verified in guarded cubical Agda \cite{veltri-vezzosi2020}, 
demonstrating that the semantics is readily mechanizable.

@@

The paper is laid out as follows:
\begin{enumerate}
\item In Section\ref{sec:GTLC} we fix our input language, a fairly
  typical gradually typed cast calculus.
\item In Section\ref{sec:concrete-term-model} we develop a first
  denotational semantics in synthetic guarded domain theory that
  satisfies adequacy and respects the equational theory, but does not
  validate graduality. We use this to introduce some of our main
  technical tools: modeling recursive types in guarded type theory and
  modeling effects using call-by-push-value.
\item In Section\ref{sec:towards-relational-model} we show where the
  New-Licata classical domain theoretic approach fails to adapt
  cleanly to the guarded setting and explore the difficulties of
  proving graduality in an intensional model.
\item In Section\ref{sec:concrete-relational-model} we describe the resulting
  model in detail. We then prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision $M \ltdyn
  N : \nat$ has the expected semantics, that $M$ errors or $M$ and $N$ have the
  same extensional behavior.
\item In Section\ref{sec:discussion} we discuss prior work on proving
  graduality, our partial mechanization of these results and discuss
  future directions for denotational semantics of gradual typing.
\end{enumerate}
